---
title: "Frequentist Calibration Curves with curveRfreq"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Frequentist Calibration Curves with curveRfreq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  eval = requireNamespace("curveRcore", quietly = TRUE)
)
```

# Ecosystem orientation

The curveR ecosystem consists of three packages that share a common
data contract defined by **curveRcore**:

* **curveRcore** — preprocessing pipelines, forward model functions,
  inverse back-calculation functions, analytical gradients, the
  `calibration_result` S3 class, and eligibility-gating logic shared
  across fitting engines.
* **curveRfreq** *(this package)* — frequentist nonlinear least-squares
  (NLS) calibration via `fit_calibration_freq()` and its multi-curve
  wrapper `fit_calibration_freq_multiplate()`.
* **curveRbayes** — Bayesian calibration via Stan; produces the same
  `calibration_result` object so downstream code is engine-agnostic.

If you need a full frequentist-vs-Bayesian comparison, including LOO-CV
model selection and posterior predictive checks, see the hub vignette
`vignette("curveR-ecosystem-overview")`. This vignette covers
**curveRfreq alone**, from raw assay data to calibrated concentrations
with per-sample precision.

---

# Data loading and preprocessing

## The `bead_assay_example` dataset

curveRfreq ships a synthetic multi-plate bead-based immunoassay dataset
that mirrors the structure expected by the fitting pipeline.

```{r load-data}
library(curveRfreq)   # also attaches curveRcore
library(ggplot2)

# Supplying package = "curveRfreq" makes data() work whether the package
# is fully installed, loaded via pkgload::load_all(), or being built by
# pkgdown — in all cases R knows exactly which data/ folder to search.
data(bead_assay_example, package = "curveRfreq")

# Top-level structure
str(bead_assay_example, max.level = 1)
```

The list contains six elements:

| Element | Description |
|---|---|
| `standards` | 60 rows × 8 cols. One row per standard well. |
| `blanks` | 24 rows × 7 cols. Four blank wells per plate. |
| `samples` | 120 rows × 13 cols. Patient samples at dilution 1:2000. |
| `curve_id_lookup` | 6 rows × 5 cols. Maps integer `curve_id` to antigen / plate metadata. |
| `response_var` | `"mfi"` — name of the response column. |
| `indep_var` | `"concentration"` — name of the independent variable column. |

The dataset spans **two antigens** (`alpha` and `beta`) × **three
replicate plates** each, giving six `curve_id` values (1–6). The
`alpha` curves were simulated with a Gompertz model; the `beta` curves
with a 5PL model, reflecting realistic between-antigen variation in
curve shape.

``` {r inspect-standards}
head(bead_assay_example$standards)
```

``` {r inspect-lookup}
bead_assay_example$curve_id_lookup
```

## Preprocessing with `curveRcore::preprocess_standards()`

**curveRfreq does not preprocess data.** All transforms — concentration
computation, prozone correction, blank handling, and log10 transforms —
must be applied upstream via `curveRcore::preprocess_standards()` (or
its component functions). This clean separation means both curveRfreq
and curveRbayes receive data on the identical fitting scale, making
results directly comparable.

`preprocess_standards()` applies these steps in order:

1. `compute_concentration()` — converts the `dilution` column to
   absolute concentration using the known undiluted standard
   concentration, optionally log10-transforming the result.
2. `correct_prozone()` — dampens the hook (prozone) effect at high
   concentrations by compressing post-peak signal toward the peak.
3. `perform_blank_operation()` — applies one of five blank strategies:
   `"ignored"` (default), `"included"` (append as extra point),
   `"subtracted"`, `"subtracted_3x"`, or `"subtracted_10x"`.
4. `compute_log_response()` — applies `log10()` to the response. Values
   ≤ 0 are floored to 1 % of the minimum positive value before
   transformation.

For the bead assay example, the undiluted standard concentration is
30 AU/mL (the highest point in the dilution series stored as
`dilution = 1/CONC`):

``` {r preprocess}
std_raw   <- bead_assay_example$standards
blank_raw <- bead_assay_example$blanks

# Preprocess one curve at a time, then stack
std_preprocessed <- do.call(rbind, lapply(
  split(std_raw, std_raw$curve_id),
  function(df) {
    cid   <- df$curve_id[1]
    blanks_for_curve <- blank_raw[blank_raw$curve_id == cid, ]
    result <- curveRcore::preprocess_standards(
      data                = df,
      antigen_settings    = list(standard_curve_concentration = 30),
      response_variable   = "mfi",
      independent_variable = "concentration",
      is_log_response     = TRUE,
      blank_data          = blanks_for_curve,
      blank_option        = "subtracted",
      is_log_independent  = TRUE,
      apply_prozone       = TRUE
    )
    result$data
  }
))

# Both response and concentration are now on the log10 scale
head(std_preprocessed[, c("curve_id", "concentration", "mfi")])
```

> **Key contract:** After `preprocess_standards()`, the column named
> `concentration` holds log10-transformed concentration, and the
> response column (`mfi` here) holds log10-transformed MFI. These are
> the scales that `fit_calibration_freq_multiplate()` expects.

---

# Fitting all curves with `fit_calibration_freq_multiplate()`

## Function signature and argument reference

`fit_calibration_freq_multiplate()` is the recommended entry point. It
splits the stacked preprocessed data by `curve_id`, calls
`fit_calibration_freq()` for each, collects results into a
`calibration_result_multiplate` object, and handles per-curve errors
gracefully.

``` {r fit}

# , cache = TRUE
# Samples stay on the RAW response scale — log-transform is applied
# internally during back-calculation.
# Note: pcov_threshold, min_dynamic_range_log10, max_rel_se, and bound_tol
# are parameters of fit_calibration_freq() only; the multiplate wrapper
# uses the single-curve defaults (20, 0.5, 5.0, 1e-4) for every curve.
samples_raw <- bead_assay_example$samples

mp <- fit_calibration_freq_multiplate(
  standards          = std_preprocessed,
  blanks             = blank_raw,
  samples            = samples_raw,
  response_var       = "mfi",
  model_names        = c("logistic4", "gompertz4"),
  is_log_response    = TRUE,    # response was log10-transformed in preprocessing
  is_log_independent = TRUE,    # concentration is on the log10 scale
  std_curve_conc     = 30,      # undiluted standard concentration, raw scale
  fixed_a            = NULL,    # estimate lower asymptote freely
  cv_x_max           = 150,     # cap pcov at 150 %
  n_grid             = 200L,    # grid resolution
  grid_min_conc      = 1e-4,    # grid lower bound, raw scale
  grid_max_conc      = NULL,    # NULL → uses std_curve_conc
  on_error           = "stop",   # DIAGNOSTIC: show actual error
  verbose            = TRUE      # DIAGNOSTIC: show fitting details
)
```

### Argument reference

| Argument | Type | Default | Meaning |
|---|---|---|---|
| `standards` | data frame | — | **Required.** Stacked preprocessed standards with `curve_id`, `concentration`, and the response column. |
| `blanks` | data frame | — | **Required.** Stacked blank data with `curve_id` and the response column, on the same fitting scale as `standards`. Stored in `result$blanks` for QA and plotting. Every `curve_id` in `standards` should have a corresponding entry. |
| `samples` | data frame or NULL | `NULL` | Stacked sample data on the **raw** (untransformed) response scale. The `dilution` column is used to recover pre-dilution concentration. |
| `response_var` | character | — | **Required.** Name of the response column. Must match what preprocessing produced. |
| `model_names` | character vector | `c("logistic4", "gompertz4")` | Which nonlinear models to fit. Any subset of `curveRcore::available_models()` is valid. `"loglogistic4"` is silently dropped when `is_log_independent = TRUE` because it is algebraically redundant with `"logistic4"` on the log scale. |
| `is_log_response` | logical | `TRUE` | Was the response log10-transformed during preprocessing? Must match the preprocessing step. |
| `is_log_independent` | logical | `TRUE` | Was concentration log10-transformed? Must match the preprocessing step. |
| `std_curve_conc` | numeric | — | **Required.** The undiluted standard concentration on the raw scale (e.g. 30 AU/mL). Used to generate the prediction grid. |
| `fixed_a` | numeric or NULL | `NULL` | If non-NULL, pins the lower asymptote `a` to this value (on the **fitting** scale, i.e. already log10-transformed if `is_log_response = TRUE`). When NULL, `a` is freely estimated. Fixing `a` is most useful when blank wells reliably define the lower asymptote. |
| `cv_x_max` | numeric | 150 | Hard cap on the percent coefficient of variation (%CV) of back-calculated concentration. Values above this threshold are set to `cv_x_max` in `$pcov` and `$pcov_rmse`. |
| `n_grid` | integer | 200 | Number of evenly spaced log10-concentration points on the prediction grid. |
| `grid_min_conc` | numeric | 1e-4 | Lower end of the prediction grid, raw scale. |
| `grid_max_conc` | numeric or NULL | `NULL` | Upper end of the prediction grid, raw scale. NULL uses `std_curve_conc`. |
| `curve_ids` | vector or NULL | `NULL` | Subset of `curve_id` values to fit. NULL fits all curves found in `standards`. |
| `on_error` | `"warn"` or `"stop"` | `"warn"` | Error handling per curve. `"warn"` stores NULL for the failed curve and continues; `"stop"` raises immediately. |
| `verbose` | logical | `FALSE` | Emit per-model convergence messages and eligibility gate results. |

Note: pcov_threshold (default 20), min_dynamic_range_log10
(default 0.5), max_rel_se (default 5.0), and bound_tol (default
1e-4) are parameters of fit_calibration_freq() and are not exposed
by the multiplate wrapper — the single-curve defaults are used for
every curve. To override them, call fit_calibration_freq() directly
for individual curves.
---

# The multi-start NLS algorithm and constraint system

Understanding the fitting engine helps interpret convergence failures
and tune the algorithm for unusual assay configurations.

## Adaptive parameter bounds

Before any fitting, `compute_model_constraints()` calls
`curveRcore::adaptive_constraint_profile()`, which inspects the
preprocessed response range and classifies the assay into one of three
**scale classes**:

| Scale class | Criterion (log response) | Typical assay |
|---|---|---|
| `"high"` | log10(MFI) max > 2.5 | Luminex / bead-based |
| `"medium"` | log10(MFI) max > 0.5 | Mid-range ELISA |
| `"low"` | log10(MFI) max ≤ 0.5 | Absorbance / OD |

The scale class drives the width of parameter bounds, which differ
systematically:

| Parameter | Role | `"high"` bounds | `"low"` bounds |
|---|---|---|---|
| `a` | Lower asymptote | `[y_min − 0.5·dr, y_min + 0.1·dr]` | same formula, narrower dr |
| `b` | Slope / steepness | `[0.1, 2.0]` | `[0.01, 5.0]` — wider to avoid singular Jacobians |
| `c` | Inflection point | ±0.5 × conc range | ±1.0 × conc range |
| `d` | Upper asymptote | `[y_min + 0.5·dr − margin, y_max + margin]` where margin = 0.5·dr | margin = 0.15·dr |
| `g` | Asymmetry (5PL) | `[0.5, 5.0]` | `[0.1, 10.0]` |

The bound on `a` is only computed when `fixed_a = NULL`; otherwise `a`
is not a free parameter and no bounds are needed.

## Multi-start Levenberg-Marquardt

`fit_ensemble_nls()` fits each model independently. For each model it:

1. **Generates starting values** from `generate_start_lists()`.
   Candidate starts are evenly spaced across the parameter bounds:
   - 3 starts for `"high"` and `"medium"` scale data.
   - 5 starts for `"low"` scale data, with the slope `b` biased toward
     the lower end of its range (squared fractional spacing) to avoid
     overshooting the flat portion of low-signal curves.

2. **Runs `minpack.lm::nlsLM()`** for each start, keeping the fit with
   the lowest AIC. Tighter tolerances (`ftol = ptol = 1e-8`, `maxiter
   = 200`) are used for low-signal data.

3. **Falls back** if all primary starts fail *and* the scale class is
   `"low"`:
   - *Fallback 1* — relaxes bounds by 50 % in each direction and
     retries `nlsLM` from the midpoint.
   - *Fallback 2* — switches to `stats::nls()` with `algorithm =
     "port"` (box-constrained trust-region), which can converge where
     Levenberg-Marquardt does not.

A model that exhausts all fallbacks is recorded with `converged =
FALSE` and is excluded from downstream eligibility assessment and
selection.

``` {r convergence-check}
# Check convergence status for curve_id 1
cr1 <- mp$plates[["1"]]

convergence <- sapply(cr1$ensemble, function(e) {
  c(model = e$model_name, converged = e$converged)
})
t(convergence)
```

---

# The eligibility-gated AIC model selection pipeline

## Why eligibility gates matter

AIC and other information criteria measure **forward-fit quality**: how
well the model predicts the response values in the training data. They
are completely blind to whether the fitted model can **back-calculate
concentration** reliably. A model whose asymptote estimate has hit a
constraint boundary, or whose covariance matrix is nearly singular,
can have a lower AIC than a better-identified competitor purely by
overfitting.

The eligibility gates intercept such models before they can be selected.
Only models that pass all applicable gates enter the AIC ranking.

## The four gates

All four gates are evaluated per model using
`curveRcore::assess_model_eligibility()`:

### Gate 1 — `at_bound`
*(Frequentist only; skipped when `constraints = NULL`)*

Checks that no parameter estimate lies within `bound_tol` (default
`1e-4`) of its lower or upper constraint bound. A boundary solution
means the optimiser could not find an interior minimum; AIC asymptotics
do not hold at the boundary, and the resulting parameter estimates are
not maximum-likelihood estimates in the usual sense.

### Gate 2 — `vcov_condition`
*(Frequentist only; skipped when `vcov(fit)` fails)*

Computes the condition number κ of `vcov(fit)`. If κ ≥ 1×10⁸, the
covariance matrix is near-singular, making standard error propagation
unreliable. This gate catches cases where two parameters are
near-collinear — for example, when the inflection point `c` lies far
outside the data range.

### Gate 3 — `rel_se`
*(Both frequentist and Bayesian)*

Checks that `SE / |estimate| < max_rel_se` (default 5.0) for every
free parameter. A relative SE larger than the estimate itself indicates
an unidentified parameter — the data do not constrain it. This gate
uses `std_error` from the NLS summary (frequentist) or `sd` from the
posterior (Bayesian).

### Gate 4 — `dynamic_range`
*(Requires `pcov_profile` and `grid_x`)*

Evaluates the precision grid (see Section 7) to locate the LLOQ and
ULOQ — defined as the grid points where `pcov` crosses `pcov_threshold`
(default 20 %) at the lower and upper ends, respectively. The gate
passes if `ULOQ − LLOQ ≥ min_dynamic_range_log10` (default 0.5 log10
units, approximately 3-fold). A model with zero quantifiable range is
operationally useless even if it converged.

## Selection logic

After assessment, `curveRcore::select_best_eligible()` implements the
following decision tree:

```
eligible_models ← {models that passed all gates}

if length(eligible_models) > 0:
    best ← eligible_models ranked by AIC (lowest wins)
    fallback ← FALSE

else:                                        # fallback path
    best ← model with widest dynamic_range_log10;
            ties broken by AIC
    fallback ← TRUE
    fallback_reason ← per-model gate failure summary
```

The fallback ensures that a result is always returned. When `fallback =
TRUE` the selection is a best-effort choice and should be treated with
caution.

---

# Reading the selection audit trail

The `$selection` component of each `calibration_result` contains the
complete audit trail.

```{r audit-trail}
sel <- cr1$selection

# Was a fallback used?
sel$fallback
sel$fallback_reason   # empty string when fallback = FALSE

# Which models passed all gates?
sel$eligible_models

# The winning model
sel$best_model_name

# AIC comparison table (always computed, even for ineligible models)
sel$weights
```

### Per-model gate results

Every model in the ensemble carries its own eligibility report inside
`ensemble[[model]]$eligibility`:

```{r gates}
for (nm in names(cr1$ensemble)) {
  elig <- cr1$ensemble[[nm]]$eligibility
  cat(sprintf("\n── %s  (eligible = %s) ──\n", nm, elig$eligible))
  print(elig$gates)
  cat(sprintf(
    "   LLOQ = %.3f  ULOQ = %.3f  DR = %.3f log10\n",
    elig$lloq, elig$uloq, elig$dynamic_range_log10
  ))
}
```

### The `aic_best` field

The selection object also exposes `$aic_best`, the model name that
would have been chosen by **pure AIC** (ignoring eligibility gates).
When `aic_best != best_model_name`, the AIC winner was demoted by a
gate failure — a useful diagnostic:

```{r aic-best}
cat("AIC winner:         ", sel$aic_best, "\n")
cat("Eligibility winner: ", sel$best_model_name, "\n")
```

---

# Per-model precision grids from `ensemble[[model]]$grid`

## Grid computation — every converged model, not just the best

A critical design point: **precision grids are computed for every
converged model**, not only the one ultimately selected. This happens
in step 6a of `fit_calibration_freq()`, before eligibility assessment:

```r
# Conceptually (internal to fit_calibration_freq):
ensemble_grids <- lapply(names(ensemble_raw), function(nm) {
  if (!ensemble_raw[[nm]]$converged) return(NULL)
  predict_grid_freq(grid = base_grid, fit = ensemble_raw[[nm]]$fit, ...)
})
```

The per-model grids are what feed the `dynamic_range` gate. After
fitting, each grid is attached to its ensemble entry as
`ensemble[[model]]$grid`. The top-level `$grid` slot of the
`calibration_result` is simply a pointer to the best-model grid — no
recomputation occurs.

## Grid columns

```{r grid-columns}
grid_best <- cr1$grid
names(grid_best)
```

| Column | Description |
|---|---|
| `x_fit` | Grid point on the fitting scale (log10 concentration, if `is_log_independent = TRUE`). |
| `log10_concentration` | Same as `x_fit` when `is_log_independent = TRUE`. |
| `predicted_response` | Forward model prediction: `ŷ = f(x, θ̂)`. On the fitting (log10) scale. |
| `ci_lower`, `ci_upper` | Delta-method 95 % CI on the **response** scale. |
| `predicted_concentration` | Back-calculated concentration from `ŷ`. On the fitting scale. |
| `se_concentration` | Delta-method SE of the back-calculated concentration. |
| `pcov` | Percent CV of back-calculated concentration (%). Capped at `cv_x_max`. |
| `pcov_rmse` | Relative RMSE (%) — includes bias of `predicted_concentration` vs `x_fit`. |
| `pcov_pass` | Logical. `TRUE` when `pcov < cv_x_max` and `pcov` is finite. |

## The `pcov` formula

For a log10-scale independent variable, `pcov` is computed from the
delta-method SE on the log10 scale:

$$
\text{pcov} = \text{SE}_{x} \times \ln(10) \times 100 \%
$$

where $\text{SE}_x^2 = \nabla_\theta x^\top \Sigma \nabla_\theta x +
\left(\frac{\partial x}{\partial y}\right)^2 \sigma_y^2$ and $\sigma_y$
is `summary(fit)$sigma` (residual SE of the response on the fitting
scale). The first term captures **parameter uncertainty** (propagated
through the inverse function via `curveRcore::make_inv_and_grad_fixed()`),
and the second captures **observation-level noise**.

## Comparing grids across models

Because every converged model has a grid, you can compare precision
profiles side by side — even for the non-selected model:

```{r compare-grids, fig.cap = "Precision profiles for both candidate models on curve_id 1."}
library(ggplot2)

grid_list <- lapply(names(cr1$ensemble), function(nm) {
  g <- cr1$ensemble[[nm]]$grid
  if (is.null(g) || !"x_fit" %in% names(g)) return(NULL)
  g <- as.data.frame(g)  # explicit copy — prevents mutation of cr1$grid
  g$model <- nm
  g
})
grid_compare <- do.call(rbind, Filter(Negate(is.null), grid_list))

if (!is.null(grid_compare) && "x_fit" %in% names(grid_compare)) {
  ggplot(grid_compare,
         aes(x = x_fit, y = pcov, colour = model)) +
    geom_line(linewidth = 0.9) +
    geom_hline(yintercept = 20, linetype = "dashed", colour = "grey40") +
    annotate("text", x = -3.8, y = 22, label = "pcov threshold (20 %)",
             hjust = 0, size = 3.2, colour = "grey40") +
    scale_colour_manual(values = c(logistic4 = "#0072B2", gompertz4 = "#D55E00")) +
    labs(x = "log10 concentration", y = "Percent CV (%)",
         title = "Precision profiles — curve_id 1",
         colour = "Model") +
    theme_bw(base_size = 12)
} else {
  message("Skipping compare-grids plot: x_fit not available in ensemble grids.")
}
```

---

# Dynamic range, LLOQ, and ULOQ

## Reading LOQ values from the eligibility report

The LLOQ and ULOQ (on the log10 concentration scale) are stored inside
`ensemble[[model]]$eligibility`:

```{r loq}
elig_best <- cr1$ensemble[[sel$best_model_name]]$eligibility

lloq_log10 <- elig_best$lloq
uloq_log10 <- elig_best$uloq
dr_log10   <- elig_best$dynamic_range_log10

cat(sprintf("LLOQ = %.3f log10 (%.4f AU/mL)\n", lloq_log10, 10^lloq_log10))
cat(sprintf("ULOQ = %.3f log10 (%.3f  AU/mL)\n", uloq_log10, 10^uloq_log10))
cat(sprintf("Dynamic range = %.3f log10 units (%.1f-fold)\n",
            dr_log10, 10^dr_log10))
```

## Visualising the quantifiable range

```{r lloq-plot, fig.cap = "Precision profile with LLOQ and ULOQ indicated. The shaded region is the quantifiable range."}
# Build a fresh plain data frame to avoid any reference or class issues
g_best <- data.frame(
  x_fit = cr1$grid[["x_fit"]],
  pcov  = cr1$grid[["pcov"]]
)

ggplot(g_best, aes(x = x_fit, y = pcov)) +
  annotate("rect",
           xmin = lloq_log10, xmax = uloq_log10,
           ymin = -Inf,       ymax = Inf,
           fill = "#009E73", alpha = 0.12) +
  geom_line(colour = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 20, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(lloq_log10, uloq_log10),
             linetype = "dotted", colour = "#009E73", linewidth = 0.8) +
  annotate("text", x = lloq_log10, y = 145, label = "LLOQ",
           hjust = -0.1, size = 3.5, colour = "#009E73") +
  annotate("text", x = uloq_log10, y = 145, label = "ULOQ",
           hjust = 1.1,  size = 3.5, colour = "#009E73") +
  coord_cartesian(ylim = c(0, 155)) +
  labs(x = "log10 concentration (AU/mL)",
       y = "Percent CV (%)",
       title = paste("Precision profile —", sel$best_model_name,
                     "— curve_id 1")) +
  theme_bw(base_size = 12)
```

> **Interpretation:** Any sample whose back-calculated concentration
> falls between LLOQ and ULOQ (shaded region) has a back-calculation
> %CV ≤ 20 %. Samples outside this range should be flagged or
> re-measured at a different dilution.

---

# Detection limits, shape-LOQ, and the second derivative profile

## Second derivative and curvature-based LOQ

In addition to the precision-profile-based LLOQ/ULOQ (derived from
`pcov` threshold crossings), every per-model grid now includes a
`d2y_dx2` column: the second derivative
$d^2(\log_{10} y) / d(\log_{10} x)^2$, computed from the existing grid
points with zero additional model evaluations.

The second derivative of a monotone increasing sigmoid has a peak
below the inflection point, a zero crossing at the inflection point,
and a valley above it.  The **shape-LLOQ** (peak) and **shape-ULOQ**
(valley) bracket the quasi-linear region of maximum assay sensitivity.

These are stored in `ensemble[[model]]$eligibility` as
`shape_lloq_log10`, `shape_uloq_log10`, and their natural-scale
and response-scale counterparts.

```{r d2y-profile, fig.cap = "Second derivative profile for the best eligible model (curve_id 1). The peak marks the shape-LLOQ and the valley marks the shape-ULOQ."}
best_grid <- as.data.frame(cr1$ensemble[[sel$best_model_name]]$grid)
best_elig <- cr1$ensemble[[sel$best_model_name]]$eligibility

if ("d2y_dx2" %in% names(best_grid)) {
  ok <- is.finite(best_grid$d2y_dx2)

  ggplot(best_grid[ok, ],
         aes(x = x_fit, y = d2y_dx2)) +
    geom_line(colour = "#0072B2", linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    {if (is.finite(best_elig$shape_lloq_log10))
      geom_vline(xintercept = best_elig$shape_lloq_log10,
                 linetype = "dotted", colour = "#D55E00", linewidth = 0.8)} +
    {if (is.finite(best_elig$shape_uloq_log10))
      geom_vline(xintercept = best_elig$shape_uloq_log10,
                 linetype = "dotted", colour = "#D55E00", linewidth = 0.8)} +
    {if (is.finite(best_elig$shape_lloq_log10))
      annotate("text", x = best_elig$shape_lloq_log10, y = max(best_grid$d2y_dx2[ok]) * 0.8,
               label = "shape-LLOQ", hjust = -0.1, colour = "#D55E00", size = 3.5)} +
    {if (is.finite(best_elig$shape_uloq_log10))
      annotate("text", x = best_elig$shape_uloq_log10, y = min(best_grid$d2y_dx2[ok]) * 0.8,
               label = "shape-ULOQ", hjust = 1.1, colour = "#D55E00", size = 3.5)} +
    labs(x = "log10 concentration (AU/mL)",
         y = expression(d^2*(log[10]*y) / d*(log[10]*x)^2),
         title = paste("Second derivative profile —", sel$best_model_name, "— curve_id 1")) +
    theme_bw(base_size = 12)
}
```

### Comparing shape-LOQ with pcov-LOQ

```{r shape-vs-pcov}
shape_vs_pcov <- data.frame(
  Metric          = c("Shape-LLOQ", "Shape-ULOQ", "pcov-LLOQ", "pcov-ULOQ"),
  log10_conc      = c(best_elig$shape_lloq_log10, best_elig$shape_uloq_log10,
                      best_elig$lloq, best_elig$uloq),
  stringsAsFactors = FALSE
)
shape_vs_pcov$conc_AU_mL <- 10^shape_vs_pcov$log10_conc

knitr::kable(shape_vs_pcov, digits = 3,
             caption = paste("Shape-LOQ vs pcov-LOQ for", sel$best_model_name, "(curve_id 1)"))
```

The shape-LOQ is purely geometric (depends only on curve shape), while
the pcov-LOQ incorporates measurement uncertainty.  The shape-LOQ
typically provides a tighter range.

## CI-based detection limits: LOD, MDC, and RDL

The `$detection_limits` element of each `calibration_result` contains
three families of metrics computed from parameter confidence intervals:

- **LODs (Limits of Detection):** the upper 97.5% CI of `a` (lower LOD)
  and the lower 2.5% CI of `d` (upper LOD), mapped to both the response
  and concentration scales.
- **MDC (Minimum Detectable Concentration):** the LOD response values
  inverted through the point-estimate curve.
- **RDL (Reliable Detection Limits):** LOD response values inverted
  through CI-modified curves (conservative = compressed `d`; liberal =
  expanded `d`).

```{r detection-limits}
dl <- cr1$detection_limits

if (!is.null(dl)) {
  dl_summary <- data.frame(
    Metric     = c("Lower LOD", "Upper LOD",
                   "MDC (lower)", "MDC (upper)",
                   "RDL (lower, conservative)", "RDL (upper, liberal)"),
    Response   = c(dl$lods$lower_lod_response,
                   dl$lods$upper_lod_response,
                   NA, NA, NA, NA),
    log10_conc = c(dl$lods$lower_lod_log10_conc,
                   dl$lods$upper_lod_log10_conc,
                   dl$mdc_rdl$mdc_lower_log10,
                   dl$mdc_rdl$mdc_upper_log10,
                   dl$mdc_rdl$rdl_lower_log10,
                   dl$mdc_rdl$rdl_upper_log10),
    conc_AU_mL = c(dl$lods$lower_lod_conc,
                   dl$lods$upper_lod_conc,
                   dl$mdc_rdl$mdc_lower_conc,
                   dl$mdc_rdl$mdc_upper_conc,
                   dl$mdc_rdl$rdl_lower_conc,
                   dl$mdc_rdl$rdl_upper_conc),
    stringsAsFactors = FALSE
  )

  knitr::kable(dl_summary, digits = 4,
               caption = paste("Detection limits for", dl$model_name,
                               "(curve_id 1, alpha =", dl$alpha, ")"))
}
```

### All QC limits on one plot

```{r qc-overlay, fig.cap = "Calibration curve (curve_id 1) with all quantification and detection limits overlaid."}
g  <- as.data.frame(cr1$grid)
dl <- cr1$detection_limits

p <- ggplot(g, aes(x = x_fit, y = predicted_response)) +
  geom_line(colour = "#0072B2", linewidth = 1.2)

# pcov-LOQ shaded region
if (is.finite(elig_best$lloq) && is.finite(elig_best$uloq)) {
  p <- p + annotate("rect",
                     xmin = elig_best$lloq, xmax = elig_best$uloq,
                     ymin = -Inf, ymax = Inf,
                     fill = "#009E73", alpha = 0.08) +
    geom_vline(xintercept = c(elig_best$lloq, elig_best$uloq),
               linetype = "dotted", colour = "#009E73", linewidth = 0.7)
}

# Shape-LOQ
if (is.finite(best_elig$shape_lloq_log10))
  p <- p + geom_vline(xintercept = best_elig$shape_lloq_log10,
                       linetype = "dashed", colour = "#D55E00", linewidth = 0.7)
if (is.finite(best_elig$shape_uloq_log10))
  p <- p + geom_vline(xintercept = best_elig$shape_uloq_log10,
                       linetype = "dashed", colour = "#D55E00", linewidth = 0.7)

# LOD horizontal lines
if (!is.null(dl) && is.finite(dl$lods$lower_lod_response))
  p <- p + geom_hline(yintercept = dl$lods$lower_lod_response,
                       colour = "darkorange", linewidth = 0.6)
if (!is.null(dl) && is.finite(dl$lods$upper_lod_response))
  p <- p + geom_hline(yintercept = dl$lods$upper_lod_response,
                       colour = "darkorange", linewidth = 0.6)

# MDC points
if (!is.null(dl) && is.finite(dl$mdc_rdl$mdc_lower_log10))
  p <- p + annotate("point",
                     x = dl$mdc_rdl$mdc_lower_log10,
                     y = dl$lods$lower_lod_response,
                     shape = 17, colour = "darkorange", size = 3)
if (!is.null(dl) && is.finite(dl$mdc_rdl$mdc_upper_log10))
  p <- p + annotate("point",
                     x = dl$mdc_rdl$mdc_upper_log10,
                     y = dl$lods$upper_lod_response,
                     shape = 17, colour = "darkorange", size = 3)

p + labs(x = "log10 concentration (AU/mL)",
         y = "log10(MFI)",
         title = paste("QC metrics overlay —", sel$best_model_name, "— curve_id 1")) +
  theme_bw(base_size = 12)
```

> **Legend:** Green shading = pcov quantifiable range (LLOQ to ULOQ).
> Orange dashed = shape-LOQ (curvature-based).  Orange horizontal lines
> = LOD response thresholds.  Orange triangles = MDC (concentration at
> LOD response).

---

# Sample back-calculation and final concentrations

## How back-calculation works

For each test sample, `predict_samples_freq()` applies the **analytical
inverse** of the best-fit model to the observed response:

1. **Log-transform** the raw response: `y_fit = log10(max(mfi, 1e-6))`.
2. **Invert the forward model** via the closed-form inverse (e.g. for
   4PL: $x = c + b \log \frac{y - a}{d - y}$) to obtain
   `predicted_concentration` on the log10 scale.
3. **Propagate uncertainty** via the delta method:
   $\text{Var}(x) = \nabla_\theta x^\top \Sigma \nabla_\theta x +
   \left(\frac{\partial x}{\partial y}\right)^2 \sigma_y^2$,
   yielding `se_concentration` and `pcov`.
4. **Recover pre-dilution concentration**:
   `final_concentration = 10^x * dilution`.

All analytical inverse and gradient functions reside in curveRcore
(`inverses.R`, `gradients.R`) and are selected via
`curveRcore::make_inv_and_grad_fixed()`, which returns a list of three
closures (`inv`, `grad`, `grad_y`) built once per model and reused
across all samples.

## Extracting sample predictions

`collect_samples()` concatenates sample prediction frames across all
curves:

```{r samples}
all_samples_pred <- collect_samples(mp)
head(all_samples_pred[, c(
  "curve_id", "sampleid", "mfi",
  "predicted_concentration", "final_concentration",
  "se_concentration", "pcov", "pcov_pass"
)])
```

### Output columns added by `predict_samples_freq()`

| Column | Description |
|---|---|
| `raw_assay_response` | Original response value (e.g. MFI) before any transform. |
| `observed_response_fit` | Response on the fitting scale (`log10(mfi)` when `is_log_response = TRUE`). |
| `predicted_log10_concentration` | Back-calculated concentration on the log10 scale. |
| `predicted_concentration` | Same as `predicted_log10_concentration` when `is_log_independent = TRUE`. |
| `final_concentration` | Pre-dilution concentration: `10^predicted_concentration × dilution`. |
| `se_concentration` | Delta-method SE of `predicted_concentration`. |
| `pcov` | %CV of back-calculated concentration. |
| `pcov_rmse` | Relative RMSE; for samples equals `pcov` (bias undefined without known truth). |
| `pcov_pass` | `TRUE` when `pcov < cv_x_max` and finite. |

## Flagging out-of-range samples

```{r flag-oor}
n_total       <- nrow(all_samples_pred)
n_pass        <- sum(all_samples_pred$pcov_pass, na.rm = TRUE)
n_out_of_range <- n_total - n_pass

cat(sprintf("Total samples : %d\n", n_total))
cat(sprintf("pcov_pass = TRUE : %d  (%.1f %%)\n",
            n_pass, 100 * n_pass / n_total))
cat(sprintf("Out-of-range    : %d  (%.1f %%)\n",
            n_out_of_range, 100 * n_out_of_range / n_total))
```

```{r sample-plot, fig.cap = "Back-calculated serum concentrations for all samples, coloured by pcov_pass status."}
# Convert predicted_concentration (log10 scale) back to natural scale
all_samples_pred$conc_AU_mL <- 10^all_samples_pred$predicted_concentration *
                                 all_samples_pred$dilution

ggplot(all_samples_pred,
       aes(x = factor(curve_id), y = log10(conc_AU_mL),
           colour = pcov_pass)) +
  geom_jitter(width = 0.25, size = 1.5, alpha = 0.8) +
  scale_colour_manual(values = c("TRUE" = "#009E73", "FALSE" = "#CC3311"),
                      labels = c("TRUE" = "Within range",
                                 "FALSE" = "Out of range")) +
  labs(x = "curve_id", y = "log10 final concentration (AU/mL)",
       colour = "pcov_pass",
       title = "Back-calculated concentrations — all curves") +
  theme_bw(base_size = 12)
```

---

# The `calibration_result` object structure

## Single-curve object

`fit_calibration_freq()` returns a `calibration_result` S3 object
(defined in curveRcore). Its components are:

```{r cr-structure}
cr1 <- mp$plates[["1"]]
str(cr1, max.level = 1)
```

| Component | Type | Contents |
|---|---|---|
| `$meta` | list | Fitting metadata: method, package, version, timestamp, `curve_id`, `response_var`, `is_log_response`, `n_standards`, `n_samples`, `pcov_threshold`. |
| `$ensemble` | named list | One entry per model name. Each entry contains `model_name`, `converged`, `parameters`, `fit_stats`, `raw_fit`, **`grid`**, and **`eligibility`**. |
| `$selection` | list | `best_model_name`, `criterion`, `fallback`, `fallback_reason`, `assessments`, `eligible_models`, `weights` (AIC table), `aic_best`. |
| `$grid` | data frame | Precision grid for the best model (pointer to `ensemble[[best]]$grid`). |
| `$samples` | data frame or NULL | Back-calculated predictions for test samples. |
| `$standards` | data frame or NULL | Preprocessed standard curve data used to fit this curve, stored for QA and plotting. |
| `$blanks` | data frame or NULL | Blank measurements for this curve, stored for QA and plotting of the zero-signal reference. |

### Accessing ensemble entries

```{r ensemble-access}
# Parameters for the best model
best_nm  <- cr1$selection$best_model_name
cr1$ensemble[[best_nm]]$parameters

# Fit statistics
cr1$ensemble[[best_nm]]$fit_stats

# Raw nlsLM object — all standard S3 methods apply
coef(cr1$ensemble[[best_nm]]$raw_fit)
```

### Accessing the precision grid for the non-selected model

Since every converged model has a grid, you can inspect precision for
the runner-up:

```{r runner-up-grid}
all_models  <- names(cr1$ensemble)
runner_up   <- setdiff(all_models, sel$best_model_name)

if (length(runner_up) > 0 &&
    cr1$ensemble[[runner_up]]$converged) {
  grid_ru <- cr1$ensemble[[runner_up]]$grid
  cat("Runner-up model:", runner_up, "\n")
  cat("Grid rows:", nrow(grid_ru), "\n")
  cat("pcov range: [",
      round(min(grid_ru$pcov, na.rm = TRUE), 1), ",",
      round(max(grid_ru$pcov, na.rm = TRUE), 1), "]\n")
}
```

## Multi-curve object

`fit_calibration_freq_multiplate()` wraps results in a
`calibration_result_multiplate` object:

```{r mp-structure}
str(mp, max.level = 1)
```

| Component | Contents |
|---|---|
| `$meta` | Global metadata: `method`, `package`, `curve_ids`, `n_curves`, `response_var`, timestamp. |
| `$plates` | Named list of `calibration_result` objects, one per `curve_id`. NULL entries indicate failed curves when `on_error = "warn"`. |

## Summary extraction helpers

Two helper functions operate on `calibration_result_multiplate` objects:

```{r summary-table}
# One row per curve: best model, parameters, fit statistics
tbl <- summary_table(mp)
print(tbl[, c("curve_id", "best_model", "converged", "aic", "a", "b", "c", "d")])
```

```{r collect-samples-2}
# All sample predictions in one data frame
samp_all <- collect_samples(mp)
cat("Total rows:", nrow(samp_all), "\n")
cat("Columns:", paste(names(samp_all), collapse = ", "), "\n")
```

---

# Summary

This vignette has walked through the complete curveRfreq workflow:

1. **Ecosystem** — curveRfreq is the frequentist fitting engine; it
   depends on curveRcore for preprocessing, model definitions, inverses,
   gradients, and the `calibration_result` class.

2. **Preprocessing** — `curveRcore::preprocess_standards()` transforms
   concentration and response to the log10 scale and handles blanks and
   prozone effects before any fitting occurs.

3. **Fitting** — `fit_calibration_freq_multiplate()` splits data by
   `curve_id` and fits an ensemble of NLS models per curve.

4. **NLS algorithm** — multi-start Levenberg-Marquardt with adaptive
   bounds; two fallback strategies for low-signal data.

5. **Eligibility gates** — four gates (`at_bound`, `vcov_condition`,
   `rel_se`, `dynamic_range`) intercept unidentified or
   boundary-constrained models before AIC ranking.

6. **Audit trail** — `$selection` exposes gate results, the AIC table,
   fallback status, and the distinction between the AIC winner and the
   eligibility-gated winner.

7. **Per-model grids** — `ensemble[[model]]$grid` is populated for
   **every converged model** and contains the full precision profile
   (`pcov`, `pcov_rmse`, `se_concentration`).

8. **LLOQ / ULOQ** — derived from the precision grid at the
   `pcov_threshold`; accessible from `ensemble[[model]]$eligibility`.
   
9. **Second derivative and shape-LOQ** — `d2y_dx2` column on every
   grid; shape-LLOQ and shape-ULOQ in `$eligibility` bracket the
   quasi-linear region of maximum sensitivity.

10. **Detection limits** — LODs, MDC, and RDL in `$detection_limits`,
    computed from asymptote parameter CIs and the analytical inverse.

11. **Sample back-calculation** — analytical inverse + delta-method
   uncertainty propagation; `final_concentration` includes the dilution
   factor; `pcov_pass` flags whether the sample falls within the
   quantifiable range.

12. **Object structure** — `calibration_result` is a standard S3 list;
    `summary_table()` and `collect_samples()` extract tidy summaries
    from multi-curve runs.
```
