---
title: "Frequentist Modeling of Immunoassay Standard Curves with the StandardCurve Class: ELISA Example"
author: ["Seamus O. Stein", "Michael Scot Zens", "Anne G. Hoen"]
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequentist Modeling of Immunoassay Standard Curves with the StandardCurve Class: ELISA Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
 
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview
This vignette demonstrates a complete workflow for fitting standard curves
for immunoassay data, including both ELISA data,
and deriving detection metrics using the `curveRfreq` package. The
workflow is encapsulated in the `StandardCurve` R6 class, which wraps
the full `curveRfreq` pipeline into a single stateful object with a
clean, verb-driven API: the class handles everything from antigen
selection through model fitting, QC metrics, error propagation, and plotting.

We walk through the complete workflow:

1. **Load and prepare data** — load the built-in `elisa_assay_example` dataset
   and construct the `StandardCurve` object with study settings and antigen constraints.
2. **Set Curve Settings** (`$set_curve_settings()`) — Resolves antigen-specific curve
   settings for the data loaded into the StandardCurve object, including lower asymptote
   handling (fix, constrained, etc.) and blank variability estimates.
3. **Fit** (`$fit()`) — fit candidate nonlinear models with the constants set in the
   object and select the best model grounded in information theory.
4. **Compare** (`$compare_models()`) — Provides a visual comparison of the converging
   models fit including an overlay, residuals vs. fitted values, the parameter estimates,
   and the AIC.
5. **Summarize** (`$summarize()`) — inspect fitted parameters and key
   QC metrics for the standard curve:
   - Inflection Point
   - Limits of Detection (LOD)
   - Minimum/Maximum Detectable Concentrations (MDC)
   - Reliable Detection Limits (RDL)
   - Curvature-based Limits of Quantification (LOQ)
6. **Plot** (`$plot()`) — visualize the final standard curve
7. **Propagate error** (`$propagate_error()`) — compute sample
   concentration predictions, measurement error, and the Precision Profile
8. **Extract** (`$get_results()`) — pull tables for downstream use

## Setup

We begin by loading the development version of the package. 

```{r setup, message = FALSE}
devtools::load_all()
```

## Load Example Data
 
The package ships with `elisa_assay_example`, a synthetic multi-plate ELISA
immunoassay dataset containing standard curves, blanks, and patient samples across
one antigen (`alpha`) run on six plates.

```{r load-data}
data(elisa_assay_example)

# Inspect the curve/plate lookup table
head(elisa_assay_example$curve_id_lookup)
```

`elisa_assay_example` is a plain list. The `StandardCurve` class expects it to contain
`$standards`, `$blanks`, `$samples`, `$response_var`, and `$indep_var`.

We choose a `curve_id` from the lookup table and use a custom filtering function (`filter_by_curve_id)` to subset the dataset of interest (`elisa_assay_example` in this vignette). This function takes the full dataset and the chosen `curve_id`, filters all relevant components, and returns a reduced dataset containing only records associated with that specific curve. The filtered dataset is then passed to the constructor.

```{r filter-function}
filter_by_curve_id <- function(loaded_data,
                               curve_id,
                               target_names = c("standards", "blanks",
                                                "samples", "curve_id_lookup"),
                               verbose = FALSE) {
  
  filtered <- loaded_data
  
  filtered$curve_id_whole_lookup <- filtered$curve_id_lookup
  filtered$whole_standards <- filtered$standards
  filtered[target_names] <- lapply(filtered[target_names], function(df) {
    
    # Skip non-data frames
    if (!is.data.frame(df)) {
      if (verbose) message("[filter_by_curve_id] Skipping non-data.frame")
      return(df)
    }
    
    # Skip empty data
    if (nrow(df) == 0) {
      if (verbose) message("[filter_by_curve_id] Empty data.frame")
      return(df)
    }
    
    # Skip if no curve_id column
    if (!"curve_id" %in% names(df)) {
      if (verbose) message("[filter_by_curve_id] No curve_id column")
      return(df)
    }
    
    # Filter
    df[as.character(df$curve_id) == as.character(curve_id), , drop = FALSE]
  })
  
  return(filtered)
}
```


```{r filter-data}
curve_id   <- 6
antigen <- elisa_assay_example$curve_id_lookup[elisa_assay_example$curve_id_lookup$curve_id == curve_id,]$antigen
curve_data <- filter_by_curve_id(elisa_assay_example, curve_id = curve_id)
```

## Configuration

- `model_names`: Candidate nonlinear models considered during the fitting algorithm.
  Model names are specified as a vector of strings limited to:
    - logistic5
    - loglogistic5
    - logistic4
    - loglogistic4
    - gompertz4

**To manually exclude a model form from consideration do not include it in the list.**

- `is_display_log_response` and `is_display_log_independent` flags control whether plotting results are displayed on log scales or not. 
 
```{r model-settings}
model_names <- c("logistic5", "loglogistic5",
                 "logistic4", "loglogistic4",
                 "gompertz4")

is_display_log_response    <- TRUE
is_display_log_independent <- TRUE
verbose                    <- TRUE
```

## Define Antigen Constraints and Study Parameters

These are specified once and passed to `$new()`. They apply to every
`$set_curve_settings()` / `$fit()` call on this object.

### Antigen Constraints

For the particular antigen in a study and experiment, constraints on the lower
asymptote of the standard curve and constraint method as well as the standard curve
concentration is set. The lower asymptote `l_asy_constraint_method` can be set to
the following:

- `default`: the default constraints built into the algorithm
- `user_defined`: uses `l_asy_min_constraint` and `l_asy_max_constraint` to constrain
  the lower asymptote. This is helpful if one wishes to fix the lower asymptote to a
  constant such as an MFI value of 0.
- `range_of_blanks`: uses the range of the blanks for the associated antigen and plate
  identifier as the lower asymptote.
- `geometric_mean_of_blanks`: constrains the lower asymptote for the associated antigen
  and plate identifier to the geometric mean of the blanks.

```{r antigen-constraints}
antigen_constraints <- data.frame(
  antigen                      = antigen,
  l_asy_min_constraint         = 0,
  l_asy_max_constraint         = 0,
  l_asy_constraint_method      = "default",
  standard_curve_concentration = 10000,
  pcov_threshold               = 15,
  stringsAsFactors             = FALSE
)

print(antigen_constraints)
```

### Study Parameters

Parameters for the entire study are set in a named list before fitting the standard
curve and influence its fit. The following are the parameters and their accepted
values and definitions. The example below uses the suggested defaults.

- **`applyProzone`**  
  Logical (`TRUE` or `FALSE`).  
  Determines whether prozone (hook effect) detection and correction is applied.  
  - `TRUE`: Detects and adjusts for potential signal suppression at low dilutions.  
  - `FALSE`: No prozone correction is performed.

- **`blank_option`**  
  Character string. Controls how blank wells are handled. Options include:  
  - `"ignored"`: Blank wells are excluded from analysis.  
  - `"subtracted"`: The geometric mean of blank responses is subtracted from observed responses.
  - `"subtracted_3x"`: 3 times the geometric mean of blank responses is subtracted from observed responses.
  - `"subtracted_10x"`: 10 times the geometric mean of blank responses is subtracted from observed responses.

  MFI values from test samples below zero are not possible (fluorescence intensity is
  physically constrained to non-negative values) and are set to zero.

- **`is_log_response`**  
  Logical (`TRUE` or `FALSE`).  
  Indicates whether the response variable (e.g., MFI or absorbance) should be
  log-transformed.  
  - `TRUE`: Fit models on the log-transformed response.  
  - `FALSE`: Use the raw response values.

- **`is_log_independent`**  
  Logical (`TRUE` or `FALSE`).  
  Specifies whether the independent variable (concentration) is log-transformed.  
  - `TRUE`: Use log-scale for the independent variable.  
  - `FALSE`: Use the original scale.

```{r study-params}
study_params <- list(
  applyProzone       = TRUE,
  blank_option       = "ignored",
  is_log_response    = TRUE,
  is_log_independent = TRUE
)
```

## Construct the StandardCurve Object

Pass the filtered data, study parameters, and constraints to `$new()`. The class
reads `response_var` and `indep_var` directly from `curve_data`.

```{r new}
sc <- StandardCurve$new(
  loaded_data                = curve_data,
  study_params               = study_params,
  antigen_constraints        = antigen_constraints,
  model_names                = model_names,
  is_display_log_response    = is_display_log_response,
  is_display_log_independent = is_display_log_independent,
  verbose                    = verbose
)

# Printing the object shows pipeline status at any time
sc
```

## Set Curve Settings

`$set_curve_settings()` receives the data in the `StandardCurve` object containing
a single antigen / plate combination and resolves antigen-specific constraints. It
resets any stale fit results automatically so re-selecting is always safe.

```{r set-curve-settings}
sc$set_curve_settings()
```

The filtered standards and antigen settings are now available for inspection if desired.

```{r set-curve-settings-inspect}
head(sc$antigen_plate$plate_standard)
sc$antigen_plate$antigen_settings
```

## Model Fitting and Selection

`$fit()` runs the entire fitting pipeline in one call across eight internal
stages: preprocessing, formula construction, constraint computation, starting
value generation, multi-start nonlinear least squares, model comparison,
AIC-based selection, and QC metric calculation.

```{r fit}
sc$fit()
```

The object now holds all intermediate results as accessible fields — nothing
is hidden.

```{r fit-internals}
# Raw candidate fits
sc$fit_summary          # AIC, BIC, RSS for each model
sc$fit_params           # parameter estimates and CIs per model
```

## Summarize

`$summarize()` computes and prints fit statistics, curve characteristics (inflection
point, LOD, MDC, RDL, LOQ), and the tidy parameter table to the console — including
the best model name, candidate model comparison (AIC, BIC, RSS), and QC metrics. A
summary of the fitted model parameters alongside their lower and upper constraints is
stored in `$best_fit$best_parameters` as a data frame, making it available for reporting
and downstream analysis.

```{r summarize}
sc$summarize()

sc$best_fit$best_model_name
sc$best_fit$best_parameters  # parameter table
sc$best_fit$best_fit_summary # QC metrics + fit stats (one row)
```

Each row corresponds to a model parameter whose definitions are as follows.

### Model Parameters

| Parameter | Description |
|-----------|-------------|
| `a` | Lower asymptote |
| `b` | Slope parameter |
| `c` | Inflection point (midpoint) |
| `d` | Upper asymptote |
| `g` | Asymmetry parameter (5-parameter models only) |

The columns include:

- `term`: Name of the model parameter
- `lower`, `upper`: Bounds for the parameter based on constraints
- `estimate`: Estimated value of the parameter
- `std_error`: Standard error of the estimate
- `statistic`: Test statistic (estimate divided by standard error)
- `p_value`: p-value of the parameter
- `curve_id`: Unique identifier for the fitted curve

The resulting `best_fit$best_fit_summary` contains one row per fitted curve and includes:

**Curve Characteristics**

| Metric | Notation | Description |
|--------|----------|-------------|
| Inflection Point | `(inflect_x, inflect_y)` | The point on the standard curve where the concavity transitions from concave up to concave down. It is the point where the assay is most sensitive to measurement errors in the measured response of the assay. |
| Limits of Detection (LOD) | `(llod, ulod)` | Lower and upper LODs are defined as the upper 97.5% confidence bound of the lower asymptote and the lower 2.5% confidence bound of the upper asymptote, respectively (Rajam et al.). Limits of Detection correspond to the y-coordinate in the legend, as they are defined on the response axis. |
| Minimum/Maximum Detectable Concentrations (MDC) | `(mindc, maxdc)` | The smallest antibody concentration that produces a signal the assay can detect above background (Rajam et al.). This corresponds to the x-coordinate of the Lower Limit of Detection in the legend, as it is on the concentration axis. |
| Reliable Detection Limits (RDL) | `(minrdl, maxrdl)` | Lower RDL: The lowest concentration at which the assay consistently produces a signal above background with 95% confidence based on the fit of the standard curve (Rajam et al.). Upper RDL: Analogously, the highest concentration at which the assay consistently produces a signal below the upper asymptote (saturation) with 95% confidence, based on the fit of the standard curve. |
| Limits of Quantification (LOQ) | `(lloq, uloq)` | Defines a region of assay response (MFI) and concentration where sample estimates have less measurement error. Limits of Quantification are derived from the local minimum and maximum of the second derivative of x given y of the standard curve (Daly et al.), (Jeanne L Sebaugh and P. D. McCray), (Sanz et al.). |

---

**Model Fit Statistics**

| Metric | Notation | Description |
|--------|----------|-------------|
| Residual Degrees of Freedom | `dfresidual` | Degrees of freedom remaining after model fitting |
| Number of Observations | `nobs` | Total number of data points used in the fit |
| R-squared | `rsquare_fit` | Goodness-of-fit measure |
| AIC | `aic` | Akaike Information Criterion for model selection |
| BIC | `bic` | Bayesian Information Criterion for model selection |
| Log-likelihood | `loglik` | Likelihood of the fitted model |
| Mean Squared Error | `mse` | Average squared difference between observed and predicted values |
| Coefficient of Variation | `cv` | Relative variability of the residuals |

---

**Model Parameters**  
Model parameter values (e.g., `a`, `b`, `c`, `d`, `g`) are included.

**Metadata**

| Field | Notation | Description |
|-------|----------|-------------|
| Model Type | `model_name` | Selected model identifier |
| Convergence Status | `status` | Indicates whether the model successfully converged |
| Standard curve formula | `formula` | Model formula used for fitting |
| Log Response Flag | `is_log_response` | Whether response variable is log-transformed |
| Log independent variable Flag | `is_log_x` | Whether independent variable (concentration) is log-transformed |
| Prozone Correction | `apply_prozone` | Indicates if prozone correction was applied |
| Background Method | `bkg_method` | Method used for blank/background handling |

## Compare Candidate Models

`$compare_models()` produces a multi-panel plot showing fitted curves,
residuals vs. fitted values, parameter estimates with confidence intervals,
and AIC scores for all converged candidate models. Use this to validate that
the AIC-selected model is sensible.

```{r compare-models, fig.width = 10, fig.height = 12, out.width = "100%"}
sc$compare_models()
```

## Propagate Measurement Error

`$propagate_error()` constructs an SE lookup table from the standards data,
retrieves the values relevant to the selected antigen, and propagates
measurement error from the standard curve to produce the precision profile
(stored in `$best_fit$sample_se`).

```{r propagate-error}
sc$propagate_error()

head(sc$best_fit$sample_se)
```

## Plot the Final Standard Curve

`$plot()` renders the selected best-fit curve with QC annotations including
the inflection point, LOD, MDC, RDL, and LOQ bands.

```{r plot, fig.width = 14, fig.height = 8, out.width = "100%"}
sc$plot()
```

Log-scale axes can be toggled without refitting:

```{r plot-linear, fig.width = 14, fig.height = 8, out.width = "100%"}
sc$plot(is_display_log_independent = FALSE, is_display_log_response = FALSE)
```

## Extract Results

`$get_results()` returns a named list of the key output tables for downstream
analysis or export.

```{r get-results}
results <- sc$get_results()

# Tables available
names(results)

# One-row QC + fit statistics summary
results$best_fit_summary

# Parameter estimates
results$best_parameters

# prediction grid from the best fitting model.
head(results$best_pred)

# Standards used for fitting
head(results$best_standard)
# Precision profile (populated after $propagate_error())
head(results$sample_se)

# Model Comparisons can be derived from the following tables. 
# This is all converged candidate fits and not solely the best fit
head(results$candidate_parameters)

# residuals vs fitted values
head(results$candidate_residuals)

```

## A more advanced example: Fitting across Multiple Plates

Since `elisa_assay_data` contains one antigen each across
six plates, we can iterate over all combinations. Call
`filter_by_curve_id()` with a new `curve_id` from the lookup table and pass the
result to `$set_data()` — it replaces the data and clears all downstream state
automatically.


```{r multi-antigen, eval = FALSE}
# Define per-antigen constraints as a named list
antigen_constraints_list <- list(
  alpha = data.frame(
    antigen                      = "alpha",
    l_asy_min_constraint         = 0,
    l_asy_max_constraint         = 0,
    l_asy_constraint_method      = "default",
    standard_curve_concentration = 10000,
    pcov_threshold               = 15,
    stringsAsFactors             = FALSE
  )
)

all_summaries <- vector("list", nrow(elisa_assay_example$curve_id_lookup))

for (i in seq_len(nrow(elisa_assay_example$curve_id_lookup))) {

  cid        <- elisa_assay_example$curve_id_lookup$curve_id[i]
  antigen    <- elisa_assay_example$curve_id_lookup$antigen[i]   # pull antigen name
  curve_data <- filter_by_curve_id(elisa_assay_example, curve_id = cid)

  # Update the StandardCurve object with new data AND new constraints
  sc$set_data(curve_data)
  sc$antigen_constraints <- antigen_constraints_list[[antigen]]  # swap constraints

  sc$set_curve_settings()$fit()$summarize()
  all_summaries[[i]] <- sc$get_results()$best_fit_summary
}

combined_summary <- do.call(rbind, all_summaries)
combined_summary
```
Key Notes:

- Pull the antigen name from the lookup table on each iteration so you know which constraint set to apply. This assumes `curve_id_lookup` has an antigen column. 

- Swap `sc$antigen_constraints` before `$set_curve_settings()` because `set_curve_settings()` reads from antigen_constraints to configure the model bounds. If you swap after, the new constraints will not take effect until the next call.

- `$set_data()` clears downstream state (fitted models, summaries, etc.) but it does not reset antigen_constraints, so the swap step is still necessary and will persist safely until overwritten on the next iteration.

- Alternatively, if the constraint differences are substantial or you want a cleaner separation, instantiate a fresh `StandardCurve` object per antigen inside the loop using `StandardCurve$new()` with the appropriate constraint data frame. This avoids any risk of stale state but is slightly more verbose.

## Quick Reference

```{r quick-ref, eval = FALSE}
# # Full single-plate workflow using the built-in example data
# data(bead_assay_example)
# 
# curve_data <- filter_by_curve_id(bead_assay_example, curve_id = 1)
# 
# sc <- StandardCurve$new(curve_data, study_params, antigen_constraints,
#                         model_names = model_names,
#                         is_display_log_response    = TRUE,
#                         is_display_log_independent = TRUE)
# 
# sc$set_curve_settings()
# sc$fit()
# sc$summarize()
# sc$compare_models()
# sc$plot()
# sc$propagate_error()
# 
# results <- sc$get_results()
# 
# # Check status at any time
# sc   # or print(sc)
```
